---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
First, load libraries
```{r, echo=FALSE}
library(dplyr)
library(lubridate)
library(data.table)
library(lattice)
```


##Loading and preprocessing the data
read activity.csv to activities variable
```{r, echo=TRUE}
activities <- read.csv('activity.csv', header = T)
```

## What is mean total number of steps taken per day?
Group and summarise activities_per_day to get table that has total steps for each day
```{r, echo=TRUE}
activities_per_day <- group_by(activities, date)
activities_per_day <- summarise(activities_per_day, total_steps = sum(steps))
```

Mean and Median value for total steps each day
```{r, echo=TRUE}
mean_value <- mean(activities_per_day$total_steps, na.rm = T)
median_value <- median(activities_per_day$total_steps, na.rm = T)
```

Plot the graph
```{r, echo=TRUE}
plot(activities_per_day$total_steps, main="Histogram of steps taken each day", 
     xlab="Date (October to November 2012)", ylab="Frequency", type="h", lwd=4, col="grey")
```


## What is the average daily activity pattern?
Group and summarise activities_per_day to get table that has average steps for each interval
```{r, echo=TRUE}
activities_per_interval <- group_by(activities, interval)
activities_per_interval <- summarise(activities_per_interval[!is.na(activities_per_interval$steps),], mean_steps = mean(steps))
```

Find out Max step and add the line in graph
```{r, echo=TRUE}
maxStepsByInterval <- max(activities_per_interval$mean_steps)
activities_per_interval_table <- data.table(activities_per_interval)
maxInterval <- activities_per_interval_table[mean_steps == maxStepsByInterval]$interval
```

Plot the graph
```{r, echo=TRUE}
plot(activities_per_interval, type="l",
     main="Average number of steps per interval across all days", 
     xlab="Interval", ylab="Average # of steps across all days", 
     lwd=2, col="grey")

abline(v=maxInterval, col="red", lwd=3)
```

## Imputing missing values
Print out value of total activities that do not have steps value
```{r, echo=TRUE}
sum(is.na(activities$steps))
```

Function to return steps. If steps is NA, return average steps of the interval instead
```{r, echo=TRUE}
getSteps = function(steps,averageSteps){
  if(is.na(steps)){
    return(averageSteps)
  }
  return(steps)
}
```

Plot 2 graphs to compare before and after filling up NA with average steps
```{r, echo=TRUE}
activities_table <- data.table(activities)
activities_per_interval_table <- data.table(activities_per_interval)
activities_table$newSteps <- mapply(getSteps,activities_table$steps, activities_per_interval_table$mean_steps)

par(mfrow=c(2,1))
plot(activities_per_day$total_steps, main="Histogram of steps taken each day", 
     xlab="Date (October to November 2012)", ylab="Frequency", type="h", lwd=4, col="grey")


activities_filled_per_day <- group_by(activities_table, date)
activities_filled_per_day <- summarise(activities_filled_per_day, total_steps = sum(newSteps))
mean_value_filled <- mean(activities_filled_per_day$total_steps, na.rm = T)
median_value_filled <- median(activities_filled_per_day$total_steps, na.rm = T)

plot(activities_filled_per_day$total_steps, main="Histogram of steps taken each day", 
     xlab="Date (October to November 2012)", ylab="Frequency", type="h", lwd=4, col="grey")
```

## Are there differences in activity patterns between weekdays and weekends?
Function to return "weekend" or "weekday" by day of the week
```{r, echo=TRUE}
day_of_the_week_type = function(x){
  if(x %in% c('1', '7')){
    return('weekend')
  }
  
  return('weekday')
}
```

Create activities table that group by weekend/weekday and interval
```{r, echo=TRUE}
activities_table <- mutate(activities_table, day_of_the_week = wday(ymd(date)))
activities_table$day_of_the_week_type <- mapply(day_of_the_week_type, activities_table$day_of_the_week)
activities_table = activities_table[, list(avg_steps = mean(newSteps, na.rm = T)), 
                                    by = list(interval, day_of_the_week_type)]
```

Plot the graph
```{r, echo=TRUE}
xyplot(avg_steps~interval | day_of_the_week_type, data = activities_table,
       type = 'l',
       xlab = 'Interval',
       ylab = 'Number of Steps',
       layout = c(1,2))
```


